---
title: "Affymetrix Processing"
subtitle: "`r paste0('Workflow Version: NF_MAAffymetrix_', params$workflow_version)`"
date: now
title-block-banner: true
format:
    html:
        code-link: true
        code-fold: true
        embed-resources: true
        toc: true
        toc-location: left
        toc-depth: 4
        number-sections: true

params:
  workflow_version: NULL
  id: NULL # str, used to name output files
  runsheet: NULL # str, path to runsheet
  biomart_attribute: NULL # str, used as a fallback value if 'Array Design REF' column is not found in the runsheet
  annotation_file_path: NULL # str, Annotation file from 'genelab_annots_link' column of GeneLab Annotations file
  ensembl_version: NULL # str, Used to determine ensembl version
  local_annotation_dir: NULL
  array_annot_path: NULL
  run_DE: 'true'

# execute: # DEBUG
#   cache: true
---

## Validate Parameters <!-- non DPPD -->
``` {r validate-parameters}
#| cache: false
#| message: false

# Ensure requisite package downloads occur in task directory
#   This is necessary since oligo attempts to install an annotations package when loading raw files.
#   This prevent permissions issues for installing and prevents side effects of processing a given dataset. 
#     (i.e. changes to more permanent package installations)
.libPaths(c(getwd(), .libPaths()))

library(dplyr) # Ensure infix operator is available, methods should still reference dplyr namespace otherwise
options(dplyr.summarise.inform = FALSE) # Don't print out '`summarise()` has grouped output by 'group'. You can override using the `.groups` argument.'
if (is.null(params$runsheet)) {
  stop("PARAMETERIZATION ERROR: Must supply runsheet path")
}

runsheet <- params$runsheet # <path/to/runsheet>

# If using custom annotation, local_annotation_dir is path to directory containing annotation file and array_annot_path is path/url to custom probe annotation file
local_annotation_dir <- params$local_annotation_dir # <path/to/annotation_dir>
array_annot_path <- params$array_annot_path # <path/to/probe_annot>

message(params)

## Set up output structure

# Output Constants
DIR_RAW_DATA <- "00-RawData"
DIR_NORMALIZED_EXPRESSION <- "01-oligo_NormExp"
DIR_DGE <- "02-limma_DGE"

dir.create(DIR_RAW_DATA)
dir.create(DIR_NORMALIZED_EXPRESSION)
dir.create(DIR_DGE)

## Save original par settings
##   Par may be temporarily changed for plotting purposes and reset once the plotting is done

original_par <- par()
options(preferRaster=TRUE) # use Raster when possible to avoid antialiasing artifacts in images

options(timeout=1000) # ensure enough time for data downloads
```

## Load Metadata and Raw Data

``` {r load-runsheet-and-annotation-table-link}
#| cache: false
#| message: false
print("Loading Runsheet...") # NON_DPPD

df_rs <- read.csv(runsheet, check.names = FALSE) %>% 
          dplyr::mutate_all(function(x) iconv(x, "latin1", "ASCII", sub="")) # Convert all characters to ascii, when not possible, remove the character

# NON_DPPD:START
print("Here is the embedded runsheet")
DT::datatable(df_rs)
# NON_DPPD:END
print("Loading Raw Data...") # NON_DPPD
local_paths <- ifelse(
  stringr::str_detect(df_rs$`Array Data File Path`, "\\.gz$"),
  stringr::str_remove(df_rs$`Array Data File Name`, "\\.gz$"),
  df_rs$`Array Data File Name`
)
df_local_paths <- data.frame(`Sample Name` = df_rs$`Sample Name`, `Local Paths` = local_paths, check.names = FALSE)
# NON_DPPD:START
print("Raw Data Location Resolved")
DT::datatable(df_local_paths)
# NON_DPPD:END

# Utility function to improve robustness of function calls
# Used to remedy intermittent internet issues during runtime 
retry_with_delay <- function(func, ...) {
  max_attempts = 5
  initial_delay = 10
  delay_increase = 30
  attempt <- 1
  current_delay <- initial_delay
  while (attempt <= max_attempts) {
    result <- tryCatch(
      expr = func(...),
      error = function(e) e
    )
    
    if (!inherits(result, "error")) {
      return(result)
    } else {
      if (attempt < max_attempts) {
        message(paste("Retry attempt", attempt, "failed for function with name <", deparse(substitute(func)) ,">. Retrying in", current_delay, "second(s)..."))
        Sys.sleep(current_delay)
        current_delay <- current_delay + delay_increase
      } else {
        stop(paste("Max retry attempts reached. Last error:", result$message))
      }
    }
    
    attempt <- attempt + 1
  }
}

# Load raw data into R object
# Retry with delay here to accomodate oligo's automatic loading of annotation packages and occasional internet related failures to load
raw_data <- retry_with_delay(
              oligo::read.celfiles,
              df_local_paths$`Local Paths`,
              sampleNames = df_local_paths$`Sample Name`# Map column names as Sample Names (instead of default filenames)
            )

print(str(raw_data))

# Summarize raw data
print("Summarized Raw Data Below") # NON_DPPD
print(paste0("Number of Arrays: ", dim(raw_data)[2]))
print(paste0("Number of Probes: ", dim(raw_data)[1]))
message(paste0("Number of Arrays: ", dim(raw_data)[2])) # NON_DPPD
message(paste0("Number of Probes: ", dim(raw_data)[1])) # NON_DPPD
# NON_DPPD:START
DT::datatable(raw_data$targets, caption = "Sample to File Mapping")
DT::datatable(head(raw_data$genes, n = 20), caption = "First 20 rows of raw data file embedded probes to genes table")
# NON_DPPD:END

annotation_file_path <- params$annotation_file_path
ensembl_version <- params$ensembl_version
```

## QA For Raw Data

### Density Plot

``` {r qa-for-raw-data--density-plot}
#| fig-cap: Density of raw intensities for each array. A lack of overlap indicates a need for normalization.
#| warning: false
#| column: screen-inset-right # Allow images to flow all the way to the right
#| fig-width: 14
#| fig-height: 8
#| fig-align: left

# Plot settings
par(
  xpd = TRUE # Ensure legend can extend past plot area
)

number_of_sets = ceiling(dim(raw_data)[2] / 30) # Set of 30 samples, used to scale plot
scale_factor = 0.2 # Default scale factor

if (max(nchar(colnames(raw_data@assayData$exprs))) > 35 & number_of_sets > 1) { # Scale more if sample names are long
  scale_factor = if_else(number_of_sets == 2, 0.4, 0.25)
}

oligo::hist(raw_data, 
            transfo=log2, # Log2 transform raw intensity values
            which=c("all"),
            nsample=10000, # Number of probes to plot
            main = "Density of raw intensities for multiple arrays")
legend("topright", legend = colnames(raw_data@assayData$exprs),
        lty = c(1,2,3,4,5), # Seems like oligo::hist cycles through these first five line types
        col = oligo::darkColors(n = ncol(raw_data)), # Ensure legend color is in sync with plot
        ncol = number_of_sets, # Set number of columns by number of sets
        cex = max(0.35, 1 + scale_factor - (number_of_sets*scale_factor)) # Reduce for each column beyond 1 with minimum of 35%
      )

# Reset par
par(original_par)
```

### Pseudo Image Plots

``` {r qa-for-raw-data--pseudoimage-plots}
#| message: false
#| warning: false # NAN can be produced due to log transformations
#| layout-ncol: 2
#| column: screen-inset-right # Allow images to flow all the way to the right
#| fig-align: left

for ( i in seq_along(1:ncol(raw_data))) {
  message(glue::glue("Drawing Psuedoimage for {colnames(raw_data)[i]}")) # NON_DPPD
  oligo::image(raw_data[,i], 
    transfo = log2,
    main = colnames(raw_data)[i]
    )
}
```

### MA Plots

``` {r report-ma-plots-approach}
#| column: screen-inset-right # Allow images to flow all the way to the right

if (inherits(raw_data, "GeneFeatureSet")) {
  print("Raw data is a GeneFeatureSet, using exprs() to access expression values and adding 0.0001 to avoid log(0)")
} else if (inherits(raw_data, "ExpressionSet") || inherits(raw_data, "ExpressionFeatureSet") || inherits(raw_data, "HTAFeatureSet")) { 
  print(paste0("Raw data is ", class(raw_data), ". Using default approach for this class for MA Plot"))
}
```

- M = Expression log-ratio (this sample vs. pseudo median reference chip)
- A = Average log-expression

``` {r qa-for-raw-data--ma-plots}
#| layout-ncol: 2
#| warning: false # NAN can be produced due to log transformations
#| column: screen-inset-right # Allow images to flow all the way to the right

if (inherits(raw_data, "GeneFeatureSet")) {
  MA_plot <- oligo::MAplot(
    exprs(raw_data) + 0.0001,
    transfo=log2,
    ylim=c(-2, 4),
    main="" # This function uses 'main' as a suffix to the sample name. Here we want just the sample name, thus here main is an empty string
  )
} else if (inherits(raw_data, "ExpressionSet") || inherits(raw_data, "ExpressionFeatureSet") || inherits(raw_data, "HTAFeatureSet")) { 
  MA_plot <- oligo::MAplot(
    raw_data,
    ylim=c(-2, 4),
    main="" # This function uses 'main' as a suffix to the sample name. Here we want just the sample name, thus here main is an empty string
  )
} else {
  stop(glue::glue("No strategy for MA plots for {class(raw_data)}"))
}
```

### Boxplots

``` {r qa-for-raw-data--boxplots}
#| warning: false # NAN can be produced due to log transformations
#| column: screen-inset-right # Allow images to flow all the way to the right
#| fig-width: 14
#| fig-height: !expr max(8, 2 + dim(raw_data)[2] * 0.2)
#| fig-align: left
max_samplename_length <- max(nchar(colnames(raw_data)))
dynamic_lefthand_margin <- max(max_samplename_length * 0.7, 10)
par(
  mar = c(8, dynamic_lefthand_margin, 8, 2) + 0.1, # mar is the margin around the plot. c(bottom, left, top, right)
  xpd = TRUE
  ) 
boxplot <- oligo::boxplot(raw_data[, rev(colnames(raw_data))], # Here we reverse column order to ensure descending order for samples in horizontal boxplot 
                          transfo=log2, # Log2 transform raw intensity values
                          which=c("all"),
                          nsample=10000, # Number of probes to plot
                          las = 1, # las specifies the orientation of the axis labels. 1 = always horizontal
                          ylab="",
                          xlab="log2 Intensity",
                          main = "Boxplot of raw intensities \nfor perfect match and mismatch probes",
                          horizontal = TRUE
                          )
title(ylab = "Sample Name", mgp = c(dynamic_lefthand_margin-2, 1, 0))
# Reset par
par(original_par)
```

## Background Correction

Approach reference: https://www.usu.edu/math/jrstevens/stat5570/1.4.Preprocess.pdf

``` {r background-correction}
#| message: false
# NON_DPPD:  RMA -> Convolution Background Correction
background_corrected_data <- raw_data %>% oligo::backgroundCorrect(method="rma")
```

## Between Array Normalization

``` {r between-array-normalization}
#| message: false
# Normalize background-corrected data using the quantile method
norm_data <- oligo::normalize(background_corrected_data, 
                              method = "quantile",
                              target = "core" # Use oligo default: core metaprobeset mappings
                              )
                              
# Summarize background-corrected and normalized data
print("Summarized Normalized Data Below") # NON_DPPD
print(paste0("Number of Arrays: ", dim(norm_data)[2]))
print(paste0("Number of Probes: ", dim(norm_data)[1]))
message(paste0("Number of Arrays: ", dim(norm_data)[2])) # NON_DPPD
message(paste0("Number of Probes: ", dim(norm_data)[1])) # NON_DPPD
# NON_DPPD:START
DT::datatable(raw_data$targets, caption = "Sample to File Mapping")
DT::datatable(head(raw_data$genes, n = 20), caption = "First 20 rows of raw data file embedded probes to genes table")
# NON_DPPD:END
```

## QA For Normalized Data

### Density Plot

``` {r qa-for-normalized-data--density-plot}
#| fig-cap: Density of normalized intensities for each array.  Compared to the raw data density plot, array densities should overlap more.
#| warning: false
#| column: screen-inset-right # Allow images to flow all the way to the right
#| fig-width: 14
#| fig-height: 8
#| fig-align: left

# Plot settings
par(
  xpd = TRUE # Ensure legend can extend past plot area
)

number_of_sets = ceiling(dim(norm_data)[2] / 30) # Set of 30 samples, used to scale plot

oligo::hist(norm_data, 
            transfo=log2, # Log2 transform normalized intensity values
            which=c("all"),
            nsample=10000, # Number of probes to plot
            main = "Density of normalized intensities for multiple arrays")
legend("topright", legend = colnames(norm_data@assayData$exprs),
        lty = c(1,2,3,4,5), # Seems like oligo::hist cycles through these first five line types
        col = oligo::darkColors(n = ncol(norm_data)), # Ensure legend color is in sync with plot
        ncol = number_of_sets, # Set number of columns by number of sets
        cex = max(0.35, 1 + scale_factor - (number_of_sets*scale_factor)) # Reduce for each column beyond 1 with minimum of 35%
      )

# Reset par
par(original_par)
```

### Pseudo Image Plots

``` {r qa-for-normalized-data--pseudoimage-plots}
#| message: false
#| warning: false # NAN can be produced due to log transformations
#| layout-ncol: 2
#| column: screen-inset-right # Allow images to flow all the way to the right
#| fig-align: left

for ( i in seq_along(1:ncol(norm_data))) {
  message(glue::glue("Drawing Psuedoimage for {colnames(norm_data)[i]}")) # NON_DPPD
  oligo::image(norm_data[,i], 
    transfo = log2,
    main = colnames(norm_data)[i]
    )
}
```

### MA Plots

- M = Expression log-ratio (this sample vs. pseudo median reference chip)
- A = Average log-expression

``` {r qa-for-normalized-data--ma-plots}
#| layout-ncol: 2
#| warning: false # NAN can be produced due to log transformations
#| column: screen-inset-right # Allow images to flow all the way to the right
#| fig-align: left

MA_plot <- oligo::MAplot(
    norm_data, 
    ylim=c(-2, 4),
    main="" # This function uses 'main' as a suffix to the sample name. Here we want just the sample name, thus here main is an empty string
)
```

### Boxplots

``` {r qa-for-normalized-data--boxplots}
#| warning: false # NAN can be produced due to log transformations
#| column: screen-inset-right # Allow images to flow all the way to the right
#| fig-width: 14
#| fig-height: !expr max(8, 2 + dim(norm_data)[2] * 0.2)
#| fig-align: left
max_samplename_length <- max(nchar(colnames(norm_data)))
dynamic_lefthand_margin <- max(max_samplename_length * 0.7, 10)
par(
  mar = c(8, dynamic_lefthand_margin, 8, 2) + 0.1, # mar is the margin around the plot. c(bottom, left, top, right)
  xpd = TRUE
  ) 
boxplot <- oligo::boxplot(norm_data[, rev(colnames(norm_data))], # Here we reverse column order to ensure descending order for samples in horizontal boxplot
                          transfo=log2, # Log2 transform normalized intensity values
                          which=c("all"),
                          nsample=10000, # Number of probes to plot
                          las = 1, # las specifies the orientation of the axis labels. 1 = always horizontal
                          ylab="",
                          xlab="log2 Intensity",
                          main = "Boxplot of normalized intensities \nfor perfect match and mismatch probes",
                          horizontal = TRUE
                          )
title(ylab = "Sample Name", mgp = c(dynamic_lefthand_margin-2, 1, 0))
# Reset par
par(original_par)
```

## Probeset Summarization

``` {r summarization}
#| message: false
# Call RMA but skip normalize and background correction since those have already been applied
probeset_level_data <- oligo::rma(norm_data, 
                                  normalize=FALSE, 
                                  background=FALSE
                                  )

# Summarize background-corrected and normalized data
print("Summarized Probeset Level Data Below") # NON_DPPD
print(paste0("Number of Arrays: ", dim(probeset_level_data)[2]))
print(paste0("Total Number of Probes Assigned To A Probeset: ", dim(oligo::getProbeInfo(probeset_level_data, target="core")['man_fsetid'])[1])) # man_fsetid means 'Manufacturer Probeset ID'. Ref: https://support.bioconductor.org/p/57191/
print(paste0("Number of Probesets: ", dim(unique(oligo::getProbeInfo(probeset_level_data, target="core")['man_fsetid']))[1])) # man_fsetid means 'Manufacturer Probeset ID'. Ref: https://support.bioconductor.org/p/57191/
message(paste0("Number of Arrays: ", dim(probeset_level_data)[2])) # NON_DPPD
message(paste0("Total Number of Probes Assigned To A Probeset: ", dim(oligo::getProbeInfo(probeset_level_data, target="core")['man_fsetid'])[1])) # NON_DPPD
message(paste0("Number of Probesets: ", dim(unique(oligo::getProbeInfo(probeset_level_data, target="core")['man_fsetid']))[1])) # NON_DPPD
# NON_DPPD:START
DT::datatable(raw_data$targets, caption = "Sample to File Mapping")
DT::datatable(head(raw_data$genes, n = 20), caption = "First 20 rows of raw data file embedded probes to genes table")
# NON_DPPD:END
```

## Probeset Annotations

### Get Probeset Annotations

``` {r retrieve-probeset-annotations}
#| message: false
shortened_organism_name <- function(long_name) {
  #' Convert organism names like 'Homo Sapiens' into 'hsapiens'
  tokens <- long_name %>% stringr::str_split(" ", simplify = TRUE)
  genus_name <- tokens[1]

  species_name <- tokens[2]

  short_name <- stringr::str_to_lower(paste0(substr(genus_name, start = 1, stop = 1), species_name))

  return(short_name)
}

get_biomart_attribute <- function(df_rs) {
  #' Returns resolved biomart attribute source from runsheet
  # NON_DPPD:START
  #' this either comes from the runsheet or as a fall back, the parameters injected during render
  #' if neither exist, an error is thrown
  # NON_DPPD:END

  # check if runsheet has 'biomart_attribute' column
  if ( !is.null(df_rs$`biomart_attribute`) ) {
    print("Using attribute name sourced from runsheet")
    # Format according to biomart needs
    formatted_value <- unique(df_rs$`biomart_attribute`) %>% 
                        stringr::str_replace_all(" ","_") %>% # Replace all spaces with underscore
                        stringr::str_to_lower() # Lower casing only
    return(formatted_value)
  } else {
    stop("ERROR: Could not find 'biomart_attribute' in runsheet")
  }
}

# Convert list of multi-mapped genes to string
list_to_unique_piped_string <- function(str_list) {
  #! convert lists into strings denoting unique elements separated by '|' characters
  #! e.g. c("GO1","GO2","GO2","G03") -> "GO1|GO2|GO3"
  return(toString(unique(str_list)) %>% stringr::str_replace_all(pattern = stringr::fixed(", "), replacement = "|"))
}

# ---- Generic Ensembl FTP mart-dump helpers --------------------------------
# Replaces live biomaRt::getBM() data queries with direct downloads of the same
# flat "mart dump" tables biomaRt itself queries under the hood, for both main
# Ensembl and Ensembl Genomes (plants/fungi/metazoa/protists/bacteria) releases.
# No retry/chunking/Sys.sleep: these are static per-release files on a stable
# FTP mirror, not a rate-limited live API, so a failed download is a real
# failure worth surfacing immediately rather than silently retrying.

resolve_mart_ftp_base <- function(division, organism, ensembl_version, ensembl_genomes_portal = NULL) {
  #' Resolve the FTP mart directory + per-species dataset prefix for either
  #' the main Ensembl release ("main") or an Ensembl Genomes division ("genomes")

  if (division == "genomes") {
    list(
      ftp_dir        = glue::glue("https://ftp.ebi.ac.uk/ensemblgenomes/pub/{ensembl_genomes_portal}/release-{ensembl_version}/mysql/{ensembl_genomes_portal}_mart_{ensembl_version}"),
      dataset_prefix = glue::glue("{organism}_eg_gene")
    )
  } else {
    list(
      ftp_dir        = glue::glue("https://ftp.ebi.ac.uk/pub/ensembl/release-{ensembl_version}/mysql/ensembl_mart_{ensembl_version}"),
      dataset_prefix = glue::glue("{organism}_gene_ensembl")
    )
  }
}

download_mart_dump <- function(url, col_names = NULL) {
  #' Download and load a single gzipped, tab-separated, headerless Ensembl "mart dump" table

  options(timeout = 300) # Can be further increased for downloading large files
  print(glue::glue("Mart dump URL: {url}"))

  temp_file <- tempfile(fileext = ".gz")
  download.file(url = url, destfile = temp_file, method = "libcurl") # libcurl needed for ftp(s) URLs

  mapping <- read.table(gzfile(temp_file), header = FALSE, sep = "\t", quote = "", comment.char = "")
  if (!is.null(col_names)) colnames(mapping)[seq_along(col_names)] <- col_names
  unlink(temp_file)
  return(mapping)
}

get_transcript_to_gene_mapping_from_ftp <- function(division, organism, ensembl_version, ensembl_genomes_portal = NULL) {
  #' Transcript -> Gene, from the "*_transcript__main.txt.gz" mart dump.
  #' Only used by the main-Ensembl probe->gene path (Probe -> Transcript -> Gene);
  #' Ensembl Genomes divisions map probe -> gene directly and never call this.
  #' Detects IDs by content instead of hardcoding column positions

  loc <- resolve_mart_ftp_base(division, organism, ensembl_version, ensembl_genomes_portal)
  url <- glue::glue("{loc$ftp_dir}/{loc$dataset_prefix}__transcript__main.txt.gz")
  raw <- retry_with_delay(download_mart_dump, url) # <- retried: large, always-expected file

  # Unversioned Ensembl gene/transcript IDs, e.g. ENSG00000210049 / ENSMUSG00000064336,
  # ENST00000387314 / ENSMUST00000082387. Anchored with no trailing ".N" to
  # exclude the separate versioned-ID columns present later in the same row.
  gene_col_idx <- which(sapply(raw, function(col) any(grepl("^ENS[A-Z]*G[0-9]+$", col))))
  transcript_col_idx <- which(sapply(raw, function(col) any(grepl("^ENS[A-Z]*T[0-9]+$", col))))

  stopifnot(
    "Could not uniquely identify gene ID column in transcript main table" = length(gene_col_idx) == 1,
    "Could not uniquely identify transcript ID column in transcript main table" = length(transcript_col_idx) == 1
  )

  raw %>%
    dplyr::transmute(
      ensembl_transcript_id = .data[[paste0("V", transcript_col_idx)]],
      ensembl_gene_id = .data[[paste0("V", gene_col_idx)]]
    ) %>%
    dplyr::distinct()
}

get_probe_to_gene_mapping_from_ftp <- function(division, organism, ensembl_version, biomart_attribute, ensembl_genomes_portal = NULL) {
  #' Probe -> Gene. The two BioMart divisions use different schemas here:
  #'   Main Ensembl ("main"): the efg dm file maps probe -> TRANSCRIPT; a
  #'     second join against the transcript__main table is required to reach gene.
  #'   Ensembl Genomes ("genomes", e.g. Ensembl Plants): the efg dm file maps
  #'     probe -> GENE directly for microarray platforms. No second join exists
  #'     here -- attempting one would silently produce wrong/empty results,
  #'     since the dm file's second column is already a gene ID, not a transcript ID.
  #' Only the attribute-specific dm file (probe_dm below) is 
  #' expected to legitimately be missing -- not every array design has a
  #' registered efg_* attribute in Ensembl BioMart. A failure there is
  #' treated as "fall back to custom annotation" (returns NULL), not an
  #' error. transcript__main.txt.gz, in contrast, is a large stable file
  #' that should always exist for a given release; a failure there is a
  #' real problem (network/transient) and is allowed to propagate as an
  #' actual error rather than being silently swallowed into a fallback.

  loc <- resolve_mart_ftp_base(division, organism, ensembl_version, ensembl_genomes_portal)
  dm_url <- glue::glue("{loc$ftp_dir}/{loc$dataset_prefix}__efg_{biomart_attribute}__dm.txt.gz")

  probe_dm <- tryCatch(
    if (division == "genomes") {
      download_mart_dump(dm_url, col_names = c("MAPID", "ensembl_gene_id", biomart_attribute))
    } else {
      download_mart_dump(dm_url, col_names = c("MAPID", "ensembl_transcript_id", biomart_attribute))
    },
    error = function(e) {
      message(glue::glue("Probe mapping file not available for attribute '{biomart_attribute}' ({e$message}); falling back to custom annotation"))
      NULL
    }
  )

  if (is.null(probe_dm)) return(NULL)

  if (division == "genomes") {
    return(probe_dm %>% dplyr::select(!!sym(biomart_attribute), ensembl_gene_id))
  }

  transcript_to_gene <- get_transcript_to_gene_mapping_from_ftp(division, organism, ensembl_version, ensembl_genomes_portal)

  probe_dm %>%
    dplyr::left_join(transcript_to_gene, by = "ensembl_transcript_id") %>%
    dplyr::select(!!sym(biomart_attribute), ensembl_gene_id)
}

# ---- Main probe annotation resolution -------------------------------------
organism <- shortened_organism_name(unique(df_rs$organism))
annot_key <- ifelse(organism %in% c("athaliana"), 'TAIR', 'ENSEMBL')
ENSEMBL_VERSION <- ensembl_version
expected_attribute_name <- get_biomart_attribute(df_rs)

if (organism %in% c("athaliana")) {
  # Unchanged from the prior implementation: single-step Probe -> Gene lookup
  # via Ensembl Genomes' own efg dm file, whose second column is already a
  # gene ID for this division (not a transcript ID needing a further join).
  
  ensembl_genomes_portal = "plants"
  print(glue::glue("Using Ensembl Genomes FTP to get probe mapping table. Portal: {ensembl_genomes_portal}, version: {ENSEMBL_VERSION}"))
  
  df_mapping <- get_probe_to_gene_mapping_from_ftp(
      division = "genomes",
      organism = organism,
      ensembl_version = ENSEMBL_VERSION,
      biomart_attribute = expected_attribute_name,
      ensembl_genomes_portal = ensembl_genomes_portal
    )

  # TAIR IDs in the mapping tables tend to be in the format 'AT1G01010.1' but the raw data has 'AT1G01010'
  # So here we remove the '.NNN' from the mapping table where .NNN is any number
  df_mapping$ensembl_gene_id <- stringr::str_replace_all(df_mapping$ensembl_gene_id, "\\.\\d+$", "")

  use_custom_annot <- FALSE
} else {
  expected_dataset_name <- glue::glue("{organism}_gene_ensembl")
  print(glue::glue("Expected dataset name: '{expected_dataset_name}'"))
  message(glue::glue("Expected dataset name: '{expected_dataset_name}'")) # NON_DPPD
  print(glue::glue("Expected attribute name: '{expected_attribute_name}'"))
  message(glue::glue("Expected attribute name: '{expected_attribute_name}'")) # NON_DPPD
  print(glue::glue("Searching for Ensembl Version: {ENSEMBL_VERSION}")) # NON_DPPD

  # Some probe_ids for affy_hta_2_0 may end in .hg.1 instead of .hg (how it is in biomaRt), leading to 0 results returned
  if (expected_attribute_name == 'affy_hta_2_0') {
    rownames(probeset_level_data) <- stringr::str_replace(rownames(probeset_level_data), '\\.hg\\.1$', '.hg')
  }

  probe_ids <- rownames(probeset_level_data)

  print(glue::glue("Using Ensembl biomart to get specific version of mapping table. Ensembl version: {ENSEMBL_VERSION}"))
  print(glue::glue("Attempting Ensembl FTP probe mapping table. Version: {ENSEMBL_VERSION}"))
  df_mapping <- get_probe_to_gene_mapping_from_ftp(
    division = "main",
    organism = organism,
    ensembl_version = ENSEMBL_VERSION,
    biomart_attribute = expected_attribute_name
  )

  if (!is.null(df_mapping)) {
    use_custom_annot <- FALSE
    # FTP has no server-side filter equivalent to getBM(filters=, values=)
    # the full mart-dump file is always downloaded in full, then filtered
    # client-side down to just this experiment's probes.
    df_mapping <- df_mapping %>% dplyr::filter(!!sym(expected_attribute_name) %in% probe_ids)
  } else {
    use_custom_annot <- TRUE
  }
}

# At this point, we have df_mapping from either the Ensembl FTP mart dumps (main or Ensembl Genomes) depending on the organism
# If no df_mapping obtained (e.g., organism not supported in biomart), use custom annotations; otherwise, merge in-house annotations to df_mapping

if (use_custom_annot) {
  expected_attribute_name <- 'ProbesetID'
  annot_type <- 'NO_CUSTOM_ANNOT'

  if (!is.null(local_annotation_dir) && !is.null(array_annot_path)) {
    probe_annot_df <- read.csv(array_annot_path, row.names=1)
    if (unique(df_rs$`biomart_attribute`) %in% row.names(probe_annot_df)) {
      annot_config <- probe_annot_df[unique(df_rs$`biomart_attribute`), ]
      annot_type <- annot_config$annot_type[[1]]
    } else {
      warning(paste0("No entry for '", unique(df_rs$`biomart_attribute`), "' in provided custom probe annotation file: ", array_annot_path))
    }
  } else {
    warning("Need to provide both local_annotation_dir and array_annot_path to use custom annotation.")
  }

  if (annot_type == '3prime-IVT') {
    unique_probe_ids <- read.csv(
      file.path(local_annotation_dir, annot_config$annot_filename[[1]]),
      skip = 13, header = TRUE, na.strings = c('NA', '---')
    )[c('Probe.Set.ID', 'Entrez.Gene', 'Gene.Symbol', 'Gene.Title', 'Ensembl', 'RefSeq.Transcript.ID', 'RefSeq.Protein.ID', 'Gene.Ontology.Biological.Process', 'Gene.Ontology.Cellular.Component', 'Gene.Ontology.Molecular.Function')]

    # Clean columns
    unique_probe_ids$Gene.Symbol <- purrr::map_chr(stringr::str_split(unique_probe_ids$Gene.Symbol, stringr::fixed(' /// ')), ~paste0(unique(.), collapse = "|")) %>% stringr::str_replace('NA', NA_character_)
    unique_probe_ids$Gene.Title <- purrr::map_chr(stringr::str_split(unique_probe_ids$Gene.Title, stringr::fixed(' /// ')), ~paste0(unique(.), collapse = "|")) %>% stringr::str_replace('NA', NA_character_)
    unique_probe_ids$Entrez.Gene <- purrr::map_chr(stringr::str_split(unique_probe_ids$Entrez.Gene, stringr::fixed(' /// ')), ~paste0(unique(.), collapse = "|")) %>% stringr::str_replace('NA', NA_character_)
    unique_probe_ids$Ensembl <- purrr::map_chr(stringr::str_split(unique_probe_ids$Ensembl, stringr::fixed(' /// ')), ~paste0(unique(.), collapse = "|")) %>% stringr::str_replace('NA', NA_character_)

    unique_probe_ids$RefSeq <- paste(unique_probe_ids$RefSeq.Transcript.ID, unique_probe_ids$RefSeq.Protein.ID)
    unique_probe_ids$RefSeq <- purrr::map_chr(stringr::str_extract_all(unique_probe_ids$RefSeq, '[A-Z]+_[\\d.]+'), ~paste0(unique(.), collapse = "|")) %>% stringr::str_replace('^$', NA_character_)

    unique_probe_ids$GO <- paste(unique_probe_ids$Gene.Ontology.Biological.Process, unique_probe_ids$Gene.Ontology.Cellular.Component, unique_probe_ids$Gene.Ontology.Molecular.Function)
    unique_probe_ids$GO <- purrr::map_chr(stringr::str_extract_all(unique_probe_ids$GO, '\\d{7}'), ~paste0('GO:', unique(.), collapse = "|")) %>% stringr::str_replace('^GO:$', NA_character_)

    unique_probe_ids <- unique_probe_ids[c('Probe.Set.ID', 'Entrez.Gene', 'Gene.Symbol', 'Gene.Title', 'Ensembl', 'RefSeq', 'GO')]
    names(unique_probe_ids) <- c('ProbesetID', 'ENTREZID', 'SYMBOL', 'GENENAME', 'ENSEMBL', 'REFSEQ', 'GOSLIM_IDS')

    unique_probe_ids$STRING_id <- NA_character_

    gene_col <- 'ENSEMBL'
    if (sum(!is.na(unique_probe_ids$ENTREZID)) > sum(!is.na(unique_probe_ids$ENSEMBL))) {
      gene_col <- 'ENTREZID'
    }
    if (sum(!is.na(unique_probe_ids$SYMBOL)) > max(sum(!is.na(unique_probe_ids$ENTREZID)), sum(!is.na(unique_probe_ids$ENSEMBL)))) {
      gene_col <- 'SYMBOL'
    }

    unique_probe_ids <- unique_probe_ids %>%
                        dplyr::mutate( 
                          count_gene_mappings = 1 + stringr::str_count(get(gene_col), stringr::fixed("|")),
                          gene_mapping_source = gene_col
                        )
  } else if (annot_type == 'custom') {
    unique_probe_ids <- read.csv(
      file.path(local_annotation_dir, annot_config$annot_filename[[1]]),
      header = TRUE, na.strings = c('NA', '')
    )
  } else {
    annot_cols <- c('ProbesetID', 'ENTREZID', 'SYMBOL', 'GENENAME', 'ENSEMBL', 'REFSEQ', 'GOSLIM_IDS', 'STRING_id', 'count_gene_mappings', 'gene_mapping_source')
    unique_probe_ids <- setNames(data.frame(matrix(NA_character_, nrow = 1, ncol = length(annot_cols))), annot_cols)
  }
} else {
  annot <- read.table(
      as.character(annotation_file_path),
      sep = "\t",
      header = TRUE,
      quote = "",
      comment.char = ""
  )

  unique_probe_ids <- df_mapping %>% 
                        # note: '!!sym(VAR)' syntax allows usage of variable 'VAR' in dplyr functions due to NSE. ref: https://dplyr.tidyverse.org/articles/programming.html # NON_DPPD
                        dplyr::mutate(dplyr::across(!!sym(expected_attribute_name), as.character)) %>% # Ensure probeset ids treated as character type
                        dplyr::group_by(!!sym(expected_attribute_name)) %>% 
                        dplyr::summarise(
                          ENSEMBL = list_to_unique_piped_string(ensembl_gene_id)
                          ) %>%
                        # Count number of ensembl IDS mapped
                        dplyr::mutate( 
                          count_gene_mappings = 1 + stringr::str_count(ENSEMBL, stringr::fixed("|")),
                          gene_mapping_source = annot_key
                        ) %>%
                        dplyr::left_join(annot, by = c("ENSEMBL" = annot_key))
}
```

``` {r reformat-merge-probe-annotations}
probeset_expression_matrix <- oligo::exprs(probeset_level_data)

probeset_expression_matrix.gene_mapped <- probeset_expression_matrix %>% 
  as.data.frame() %>%
  tibble::rownames_to_column(var = "ProbesetID") %>% # Ensure rownames (probeset IDs) can be used as join key
  dplyr::left_join(unique_probe_ids, by = c("ProbesetID" = expected_attribute_name ) ) %>%
  dplyr::mutate( count_gene_mappings := ifelse(is.na(count_gene_mappings), 0, count_gene_mappings) ) %>%
  dplyr::mutate( gene_mapping_source := unique(unique_probe_ids$gene_mapping_source) )
```

### Summarize Gene Mapping

``` {r summarize-remapping-vs-original-mapping}
#| message: false
# Pie Chart with Percentages
slices <- c(
    'Unique Mapping' = nrow(probeset_expression_matrix.gene_mapped %>% dplyr::filter(count_gene_mappings == 1) %>% dplyr::distinct(ProbesetID)), 
    'Multi Mapping' = nrow(probeset_expression_matrix.gene_mapped %>% dplyr::filter(count_gene_mappings > 1) %>% dplyr::distinct(ProbesetID)), 
    'No Mapping' = nrow(probeset_expression_matrix.gene_mapped %>% dplyr::filter(count_gene_mappings == 0) %>% dplyr::distinct(ProbesetID))
)
pct <- round(slices/sum(slices)*100)
chart_names <- names(slices)
chart_names <- glue::glue("{names(slices)} ({slices})") # add count to labels
chart_names <- paste(chart_names, pct) # add percents to labels
chart_names <- paste(chart_names,"%",sep="") # ad % to labels
pie(slices,labels = chart_names, col=rainbow(length(slices)),
    main=glue::glue("Mapping to Primary Keytype\n {nrow(probeset_expression_matrix.gene_mapped %>% dplyr::distinct(ProbesetID))} Total Unique Probesets")
    )

print(glue::glue("Unique Mapping Count: {slices[['Unique Mapping']]}"))
message(glue::glue("Unique Mapping Count: {slices[['Unique Mapping']]}")) # NON_DPPD
```

### Generate Annotated Raw and Normalized Expression Tables

```{r save-tables}
## Reorder columns before saving to file
ANNOTATIONS_COLUMN_ORDER = c(
  annot_key,
  "SYMBOL",
  "GENENAME",
  "REFSEQ",
  "ENTREZID",
  "STRING_id",
  "GOSLIM_IDS"
)

SAMPLE_COLUMN_ORDER <- df_rs$`Sample Name`

probeset_expression_matrix.gene_mapped <- probeset_expression_matrix.gene_mapped %>% dplyr::rename( !!annot_key := ENSEMBL )

## Output column subset file with just normalized probeset level expression values
write.csv(
  probeset_expression_matrix.gene_mapped[c(
  ANNOTATIONS_COLUMN_ORDER,
  "ProbesetID",
  "count_gene_mappings",
  "gene_mapping_source",
  SAMPLE_COLUMN_ORDER)
  ], file.path(DIR_NORMALIZED_EXPRESSION, "normalized_expression_probeset_GLmicroarray.csv"), row.names = FALSE)

## Determine column order for probe level tables

PROBE_INFO_COLUMN_ORDER = c(
  "ProbesetID",
  "ProbeID",
  "count_gene_mappings",
  "gene_mapping_source"
)

FINAL_COLUMN_ORDER <- c(
  ANNOTATIONS_COLUMN_ORDER, 
  PROBE_INFO_COLUMN_ORDER, 
  SAMPLE_COLUMN_ORDER
)

## Generate raw intensity matrix that includes annotations

background_corrected_data_annotated <- oligo::exprs(background_corrected_data) %>% 
  as.data.frame() %>%
  tibble::rownames_to_column(var = "fid") %>% # Ensure rownames (probeset IDs) can be used as join key
  dplyr::mutate(dplyr::across(fid, as.integer)) %>% # Ensure fid is integer type, consistent with getProbeInfo typing
  dplyr::right_join(oligo::getProbeInfo(background_corrected_data), by = "fid") %>% # Add 'man_fsetid' via mapping based on fid
  dplyr::rename( ProbesetID = man_fsetid ) %>% # Rename from getProbeInfo name to ProbesetID
  dplyr::rename( ProbeID = fid ) %>% # Rename from getProbeInfo name to ProbeID
  dplyr::left_join(unique_probe_ids, by = c("ProbesetID" = expected_attribute_name ) ) %>% # Join with biomaRt ENSEMBL mappings
  dplyr::mutate( count_gene_mappings := ifelse(is.na(count_gene_mappings), 0, count_gene_mappings) ) %>% # Convert NA mapping to 0
  dplyr::mutate( gene_mapping_source := unique(unique_probe_ids$gene_mapping_source) ) %>%
  dplyr::rename( !!annot_key := ENSEMBL )

## Perform reordering
background_corrected_data_annotated <- background_corrected_data_annotated %>% 
  dplyr::relocate(dplyr::all_of(FINAL_COLUMN_ORDER))

write.csv(background_corrected_data_annotated, file.path(DIR_RAW_DATA, "raw_intensities_probe_GLmicroarray.csv"), row.names = FALSE)

## Generate normalized expression matrix that includes annotations
norm_data_matrix_annotated <- oligo::exprs(norm_data) %>% 
  as.data.frame() %>%
  tibble::rownames_to_column(var = "fid") %>% # Ensure rownames (probeset IDs) can be used as join key
  dplyr::mutate(dplyr::across(fid, as.integer)) %>% # Ensure fid is integer type, consistent with getProbeInfo typing
  dplyr::right_join(oligo::getProbeInfo(norm_data), by = "fid") %>% # Add 'man_fsetid' via mapping based on fid
  dplyr::rename( ProbesetID = man_fsetid ) %>% # Rename from getProbeInfo name to ProbesetID
  dplyr::rename( ProbeID = fid ) %>% # Rename from getProbeInfo name to ProbeID
  dplyr::left_join(unique_probe_ids, by = c("ProbesetID" = expected_attribute_name ) ) %>%
  dplyr::mutate( count_gene_mappings := ifelse(is.na(count_gene_mappings), 0, count_gene_mappings) ) %>% # Convert NA mapping to 0
  dplyr::mutate( gene_mapping_source := unique(unique_probe_ids$gene_mapping_source) ) %>%
  dplyr::rename( !!annot_key := ENSEMBL ) 

norm_data_matrix_annotated <- norm_data_matrix_annotated %>% 
  dplyr::relocate(dplyr::all_of(FINAL_COLUMN_ORDER))

write.csv(norm_data_matrix_annotated, file.path(DIR_NORMALIZED_EXPRESSION, "normalized_intensities_probe_GLmicroarray.csv"), row.names = FALSE)
```

## Perform Probeset Differential Expression (DE)

### Generate Design Matrix

``` {r generate-design-matrix}
#| include: !expr params$run_DE
#| eval: !expr params$run_DE

runsheet_to_design_matrix <- function(runsheet_path) {
    df <- read.csv(runsheet, check.names = FALSE) %>% 
              dplyr::mutate_all(function(x) iconv(x, "latin1", "ASCII", sub="")) # Convert all characters to ascii, when not possible, remove the character    # get only Factor Value columns
    factors = as.data.frame(df[,grep("Factor.Value", colnames(df), ignore.case=TRUE)])
    colnames(factors) = paste("factor",1:dim(factors)[2], sep= "_")
    
    # Load metadata from runsheet csv file
    compare_csv = data.frame(sample_id = df[,c("Sample Name")], factors)

    # Create data frame containing all samples and respective factors
    study <- as.data.frame(compare_csv[,2:dim(compare_csv)[2]])
    colnames(study) <- colnames(compare_csv)[2:dim(compare_csv)[2]]
    rownames(study) <- compare_csv[,1] 
    
    # Format groups and indicate the group that each sample belongs to
    if (dim(study)[2] >= 2){
        group<-apply(study,1,paste,collapse = " & ") # concatenate multiple factors into one condition per sample
    } else{
        group<-study[,1]
    }
    group_names <- paste0("(",group,")",sep = "") # human readable group names
    group <- sub("^BLOCKER_", "",  make.names(paste0("BLOCKER_", group))) # group naming compatible with R models, this maintains the default behaviour of make.names with the exception that 'X' is never prepended to group namesnames(group) <- group_names
    names(group) <- group_names

    # Format contrasts table, defining pairwise comparisons for all groups
    contrast.names <- combn(levels(factor(names(group))),2) # generate matrix of pairwise group combinations for comparison
    contrasts <- apply(contrast.names, MARGIN=2, function(col) sub("^BLOCKER_", "",  make.names(paste0("BLOCKER_", stringr::str_sub(col, 2, -2)))))
    contrast.names <- c(paste(contrast.names[1,],contrast.names[2,],sep = "v"),paste(contrast.names[2,],contrast.names[1,],sep = "v")) # format combinations for output table files names
    contrasts <- cbind(contrasts,contrasts[c(2,1),])
    colnames(contrasts) <- contrast.names
    sampleTable <- data.frame(condition=factor(group))
    rownames(sampleTable) <- df[,c("Sample Name")]

    condition <- sampleTable[,'condition']
    names_mapping <- as.data.frame(cbind(safe_name = as.character(condition), original_name = group_names))

    design <- model.matrix(~ 0 + condition)
    design_data <- list( matrix = design, mapping = names_mapping, groups = as.data.frame( cbind(sample = df[,c("Sample Name")], group = group_names) ), contrasts = contrasts )
    return(design_data)
}


# Loading metadata from runsheet csv file
design_data <- runsheet_to_design_matrix(runsheet)
design <- design_data$matrix

# Write SampleTable.csv and contrasts.csv file
write.csv(design_data$groups, file.path(DIR_DGE, "SampleTable_GLmicroarray.csv"), row.names = FALSE)
write.csv(design_data$contrasts, file.path(DIR_DGE, "contrasts_GLmicroarray.csv"))
```

### Perform Individual Probeset Level DE

``` {r perform-probeset-differential-expression}
#| include: !expr params$run_DE
#| eval: !expr params$run_DE

lm_fit_pairwise <- function(norm_data, design) {
    #' Perform all pairwise comparisons

    #' Approach based on limma manual section 17.4 (version 3.52.4)

    fit <- limma::lmFit(norm_data, design)

    # Create Contrast Model
    fit.groups <- colnames(fit$design)[which(fit$assign == 1)]
    combos <- combn(fit.groups,2)
    contrasts<-c(paste(combos[1,],combos[2,],sep = "-"),paste(combos[2,],combos[1,],sep = "-")) # format combinations for limma:makeContrasts
    cont.matrix <- limma::makeContrasts(contrasts=contrasts,levels=design)
    contrast.fit <- limma::contrasts.fit(fit, cont.matrix)

    contrast.fit <- limma::eBayes(contrast.fit,trend=TRUE,robust=TRUE)
    return(contrast.fit)
}

# Calculate results
res <- lm_fit_pairwise(probeset_level_data, design)
DT::datatable(limma::topTable(res)) # NON_DPPD

# Print DE table, without filtering
limma::write.fit(res, adjust = 'BH', 
                file = "INTERIM.csv",
                row.names = FALSE,
                quote = TRUE,
                sep = ",")
```

### Add Annotation and Stats Columns and Format DE Table

``` {r save-de-table}
#| message: false
#| include: !expr params$run_DE
#| eval: !expr params$run_DE
## Reformat Table for consistency across DE analyses tables within GeneLab ##

# Read in DE table 
df_interim <- read.csv("INTERIM.csv")

# Bind columns from gene mapped expression table
df_interim <- df_interim %>% 
  dplyr::bind_cols(probeset_expression_matrix.gene_mapped)

# Reformat column names
reformat_names <- function(colname, group_name_mapping) {
  # NON_DPPD:START
  #! Converts from:
  #!    "P.value.adj.conditionWild.Type...Space.Flight...1st.generation.conditionWild.Type...Ground.Control...4th.generation"
  #! to something like:
  #! "Adj.p.value(Wild Type & Space Flight & 1st generation)v(Wild Type & Ground Control & 4th generation)"
  #! Since two groups are expected to be replace, ensure replacements happen in pairs

  # Remove 'condition' from group names
  ## This was introduced while creating design matrix
  # Rename other columns for consistency across genomics related DE outputs
  # NON_DPPD:END
  new_colname <- colname  %>% 
                  stringr::str_replace(pattern = "^P.value.adj.condition", replacement = "Adj.p.value_") %>%
                  stringr::str_replace(pattern = "^P.value.condition", replacement = "P.value_") %>%
                  stringr::str_replace(pattern = "^Coef.condition", replacement = "Log2fc_") %>% # This is the Log2FC as per: https://rdrr.io/bioc/limma/man/writefit.html
                  stringr::str_replace(pattern = "^t.condition", replacement = "T.stat_") %>%
                  stringr::str_replace(pattern = ".condition", replacement = "v")
  
  # remap to group names before make.names was applied
  unique_group_name_mapping <- unique(group_name_mapping) %>% arrange(-nchar(safe_name))
  for ( i in seq(nrow(unique_group_name_mapping)) ) {
    safe_name <- unique_group_name_mapping[i,]$safe_name
    original_name <- unique_group_name_mapping[i,]$original_name
    new_colname <- new_colname %>% stringr::str_replace(pattern = stringr::fixed(safe_name), replacement = original_name)
  }

  return(new_colname)
}

df_interim <- df_interim %>% dplyr::rename_with(reformat_names, .cols = matches('\\.condition'), group_name_mapping = design_data$mapping)


## Add Group Wise Statistics ##

# Group mean and standard deviations for normalized expression values are computed and added to the table

unique_groups <- unique(design_data$group$group)
for ( i in seq_along(unique_groups) ) {
  current_group <- unique_groups[i]
  current_samples <- design_data$group %>%
                      dplyr::filter(group == current_group) %>%
                      dplyr::pull(sample) %>%
                      sort()
                    
  print(glue::glue("Computing mean and standard deviation for Group {i} of {length(unique_groups)}"))
  print(glue::glue("Group: {current_group}"))
  print(glue::glue("Samples in Group: '{toString(current_samples)}'"))
  # NON_DPPD:START
  message(glue::glue("Computing mean and standard deviation for Group {i} of {length(unique_groups)}"))
  message(glue::glue("Group: {current_group}"))
  message(glue::glue("Samples in Group: '{toString(current_samples)}'"))
  # NON_DPPD:END
  
  df_interim <- df_interim %>% 
    dplyr::mutate( 
      "Group.Mean_{current_group}" := rowMeans(dplyr::select(., all_of(current_samples))),
      "Group.Stdev_{current_group}" := matrixStats::rowSds(as.matrix(dplyr::select(., all_of(current_samples)))),
      ) %>% 
    dplyr::ungroup() %>%
    as.data.frame()
}

# NON_DPPD:START
## Compute all sample mean and standard deviation
message(glue::glue("Computing mean and standard deviation for all samples"))
# NON_DPPD:END
df_interim <- df_interim %>% 
  dplyr::mutate( 
    "All.mean" := rowMeans(dplyr::select(., all_of(SAMPLE_COLUMN_ORDER))),
    "All.stdev" := matrixStats::rowSds(as.matrix(dplyr::select(., all_of(SAMPLE_COLUMN_ORDER)))),
    ) %>% 
  dplyr::ungroup() %>%
  as.data.frame()

print("Remove extra columns from final table")

# These columns are data mapped to column PROBEID as per the original Manufacturer and can be linked as needed
colnames_to_remove = c(
  "AveExpr" # Replaced by 'All.mean' column
)

df_interim <- df_interim %>% dplyr::select(-any_of(colnames_to_remove))

PROBE_INFO_COLUMN_ORDER = c(
  "ProbesetID",
  "count_gene_mappings",
  "gene_mapping_source"
)

generate_prefixed_column_order <- function(subjects, prefixes) {
  #' Return a vector of columns based on subject and given prefixes
  #'  Used for both contrasts and groups column name generation
  
  # Track order of columns
  final_order = c()

  # For each contrast
  for (subject in subjects) {
    # Generate column names for each prefix and append to final_order
    for (prefix in prefixes) {
      final_order <- append(final_order, glue::glue("{prefix}{subject}"))
    }
  }
  return(final_order)
}
STAT_COLUMNS_ORDER <- generate_prefixed_column_order(
  subjects = colnames(design_data$contrasts),
  prefixes = c(
    "Log2fc_",
    "T.stat_",
    "P.value_",
    "Adj.p.value_"
    )
  )
ALL_SAMPLE_STATS_COLUMNS_ORDER <- c(
  "All.mean",
  "All.stdev",
  "F",
  "F.p.value"
)

GROUP_MEAN_STDEV_COLUMNS_ORDER <- generate_prefixed_column_order(
  subjects = unique(design_data$groups$group),
  prefixes = c(
    "Group.Mean_",
    "Group.Stdev_"
  )
)

FINAL_COLUMN_ORDER <- c(
  ANNOTATIONS_COLUMN_ORDER, 
  PROBE_INFO_COLUMN_ORDER, 
  SAMPLE_COLUMN_ORDER, 
  STAT_COLUMNS_ORDER, 
  ALL_SAMPLE_STATS_COLUMNS_ORDER, 
  GROUP_MEAN_STDEV_COLUMNS_ORDER
)

## Assert final column order includes all columns from original table
if (!setequal(FINAL_COLUMN_ORDER, colnames(df_interim))) {
  write.csv(FINAL_COLUMN_ORDER, "FINAL_COLUMN_ORDER.csv")
  NOT_IN_DF_INTERIM <- paste(setdiff(FINAL_COLUMN_ORDER, colnames(df_interim)), collapse = ":::")
  NOT_IN_FINAL_COLUMN_ORDER <- paste(setdiff(colnames(df_interim), FINAL_COLUMN_ORDER), collapse = ":::")
  stop(glue::glue("Column reordering attempt resulted in different sets of columns than original. Names unique to 'df_interim': {NOT_IN_FINAL_COLUMN_ORDER}. Names unique to 'FINAL_COLUMN_ORDER': {NOT_IN_DF_INTERIM}."))
}

## Perform reordering
df_interim <- df_interim %>% dplyr::relocate(dplyr::all_of(FINAL_COLUMN_ORDER))

# Save to file
write.csv(df_interim, file.path(DIR_DGE, "differential_expression_GLmicroarray.csv"), row.names = FALSE)
```

## Version Reporting <!-- non DPPD -->

```{r version-reporting}
get_versions <- function() {
  clean_url_field <- function(url_vector) {
    # URL field can include multiple entries and newline characters
    #   This helper function extracts just the first url
    # Handle empty fields, populate downstream
    if (is.null(url_vector)) {  
      return("NO URLS ENCODED")
    }
    tryCatch(
    {return(
        (url_vector %>% 
          stringr::str_split(pattern = ",") %>% # Often split on commas
          dplyr::first() %>% # Get first token after comma split
          stringr::str_split(pattern = " ") %>% # Sometimes just spaces to split, e.g. URL: https://github.com/jeroen/curl (devel) https://curl.se/libcurl/  \n(upstream)
          dplyr::first() %>% # Get first token after space split
          stringr::str_replace_all(pattern = "\n", replacement = "") # Never allow newlines, hopefully unlikely after prior steps to isolate first url token
        )[[1]]
      )},
    error = function(cond) {
            print(url_vector)
            stop(cond)
        }
    )
 
  }
  # Note: newlines seem duplicated here as 'glue' trims the first and last newline if they exist
  session_info <- sessionInfo()
  # start with just R version
  versions_buffer <- glue::glue_collapse(c(
    glue::glue("- name: R"),
    glue::glue("  version: {session_info[['R.version']][['major']]}.{session_info[['R.version']][['minor']]}"),
    glue::glue("  homepage: https://www.r-project.org/"),
    glue::glue("  workflow task: PROCESS_AFFYMETRIX")
    ), sep = "\n") 
  # Add Bioconductor explicitly as a transitive dependency not captured by sessionInfo()
  versions_buffer <- glue::glue_collapse(c(
    versions_buffer,
    glue::glue("- name: Bioconductor"),
    glue::glue("  version: {packageVersion('BiocVersion')}"),
    glue::glue("  homepage: https://bioconductor.org"),
    glue::glue("  workflow task: PROCESS_AFFYMETRIX")
    ), sep = "\n")
  # Get 'other attached packages'
  for (software in session_info[["otherPkgs"]]) {
    versions_buffer <- glue::glue_collapse(c(
      versions_buffer,
      glue::glue("- name: {software[['Package']]}"),
      glue::glue("  version: {software[['Version']]}"),
      glue::glue("  homepage: {clean_url_field(software[['URL']])}"),
      glue::glue("  workflow task: PROCESS_AFFYMETRIX")
      ), sep = "\n") 
  }
  # Get 'loaded via a namespace (and not attached):'
  for (software in session_info[["loadedOnly"]]) {
    versions_buffer <- glue::glue_collapse(c(
      versions_buffer,
      glue::glue("- name: {software[['Package']]}"),
      glue::glue("  version: {software[['Version']]}"),
      glue::glue("  homepage: {clean_url_field(software[['URL']])}"),
      glue::glue("  workflow task: PROCESS_AFFYMETRIX")
      ), sep = "\n") 
  }
  return(versions_buffer)
}

versions_buffer <- get_versions()

## Note Libraries that were NOT used during processing
if (!grepl("purrr", versions_buffer)) {
  versions_buffer <- glue::glue_collapse(c(
    versions_buffer,
    glue::glue("- name: purrr"),
    glue::glue("  version: (Not used for this dataset)"),
    glue::glue("  homepage: https://purrr.tidyverse.org/"),
    glue::glue("  workflow task: PROCESS_AFFYMETRIX")
    ), sep = "\n")
}

## Log same info into versions.txt file
version_output_fn <- "versions.yml"
cat(versions_buffer, file = version_output_fn, append = TRUE, sep = "\n")
## Print for report
print("Session Info below: ")
print(sessionInfo())
```